---
title       : Mortgage Primer
subtitle    : The mortgage analysis tool
author      : Adeyemi Arogunmati
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r init, include=FALSE}
suppressWarnings(suppressMessages(library(lubridate)))
suppressWarnings(suppressMessages(library(ggplot2)))
```
    
## Introduction

1. Mortgage Primer is a simple mortgage analysis tool
2. It is hosted on shinyapps.io server at  <https://adeyemiarogunmati.shinyapps.io/Primer/>
3. It is designed to be used for multiple purposes
    - simple mortgage monthly payment calculation
    - loan balance calculation after n months
    - loan term required to produce monthly payment given balance
    - assessing balance reduction with or without additional payments
4. It is user friendly and visually appealing

--- .class #id 

## Example Mortgage Calculation

- Mortgage Primer (MP) functions are defined on the shiny server side

- These functions are hidden from the user

- To calculate mortgage payments, MP calls its **calcMortgagePayment** function

```{r mortgage_payment, echo=FALSE}
calcMortgagePayment <- function(apr,loan,term) {
    c <- apr/100/12   # APR per month
    n <- term*12      # loan term in months
    
    numerator   <- loan * c * (1 + c)^n
    denominator <- (1 + c)^n - 1
    
    P <- round(numerator/denominator,2)
    
    P
}
```

- An example function call is shown below:

```{r run_mortgage_calculation}
apr  <- 3.375
loan <- 340000
term <- 30
calcMortgagePayment (apr,loan,term)
```

--- .class #id 

## Example Loan Payment Assessment

- To assess loan balance reduction with yearly additional payments, MP calls its **calcAmortizationScheduleWithYearlyPrincipal** function


```{r loan_balance, echo=FALSE}
calcAmortizationScheduleWithYearlyPrincipal <- function(apr,loan,term,addPrincip,monthBegin=0) {
    n <- term*12      # loan term in months
    yp <- 0 # mp is yearly principal
    balance <- loan
    P <- calcMortgagePayment(apr,loan,term)
    
    
    outputList <- list()
    
    cumInterest  <- 0
    cumPrincipal <- 0
    
    for (i in seq(1,n)) {
        
        if ( i >= monthBegin & i%%12 == 1 ) yp <- addPrincip
        else yp <- 0
        
        I <- calcInterestPaymentOnBalance (apr,balance)
        
        if (P-I < balance & P+yp-I > balance) {
            yp <- balance - (P-I)
        }
        if (P-I >= balance) {
            P <- balance + I
            yp <- 0
        }
        
        I  <- round(I,2)
        P  <- round(P,2)
        yp <- round(yp,2)
        balance <- round(balance,2)
        
        balance0 <- balance
        balance <- balance + I - (P + yp)
        
        cumInterest  <- cumInterest  + I
        cumPrincipal <- cumPrincipal + P+yp-I
        
        outputList[[i]] <- c(balance0,i,I,P-I,P,yp,P+yp,cumInterest,cumPrincipal)
        
        if (balance^2 < 0.001) break
    }
    
    output <- as.data.frame(do.call("rbind",outputList))
    names(output) <- c("Principal.Balance","Month.Number","Interest.Payment",
                       "Principal.Payment","Mortgage.Payment","Additional.Payment",
                       "Total.Payment","Cummulative.Interest","Cummulative.Principal")
    
    output
}

addDateColumnToSchedule <- function(amortizationSchedule,startDate) {
    
    startDate <- as.Date(startDate)
    dateVec <- startDate %m+% months(seq(0,nrow(amortizationSchedule)-1))
    
    newSchedule <- cbind(dateVec,amortizationSchedule)
    
    colnames(newSchedule)[1] <- "dates"
    
    newSchedule
}

calcInterestPaymentOnBalance <- function(apr,balance) {
    c <- apr/100/12   # APR per month
    I <- balance*c
    I
}
```

- For the previous example, to assess loan figures with yearly additional payment of $10,000, MP calls:

```{r loan_balance_plot1, echo=FALSE}
startDate <- as.Date("2012-08-31")
```

```{r loan_balance_plot2}
d <- calcAmortizationScheduleWithYearlyPrincipal(apr,loan,term,10000)
```

```{r loan_balance_plot3, echo=FALSE, fig.height=4.5, fig.width=10}
d <- addDateColumnToSchedule(d,startDate)
g <- ggplot(d,aes(x = dates, y = Principal.Balance/1000, color="Loan Balance")) + 
     geom_line() + 
     geom_line(aes(x = dates, y = Cummulative.Principal/1000, color="Principal Paid")) +
     geom_line(aes(x = dates, y = Cummulative.Interest/1000, color="Interest Paid")) +
     labs(x = "Date",y = "Amount (x $1000)",title = "Amount over time") + 
     theme(plot.background = element_rect(fill = "transparent",colour = NA))

plot(g)
```

--- .class #id 

## Conclusions

- Mortgage Primer (MP) is an easy to use mortgage analysis tool
- Its functions are defined in files hosted on the shiny.io server
- Its plots are created with visually appealing ggplot calls
- MP can be accessed by clicking this [web link] (https://adeyemiarogunmati.shinyapps.io/Primer/)

